Association between trauma severity models and opportunities for improvement:

### 2022-09-16 studyprogressreport.Rmd
study progress report ready for feedback

---
title: "Association between trauma severity models and opportunities for improvement: A retrospective cohort study"
author: "Victoria Witthoff"
output:
  pdf_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

Abbreviations
=============
* ASA-PS - American Society of Anesthesiology Physical Status
* KUH - Karolinska University Hospital
* NORMIT - Norwegian Survival Prediction Model in Trauma
* OFI - Opportunities for improvement
* Ps - Probability of survival
* TCQD - Karolinska University Hospital Trauma Care Quality Database
* TRISS - Trauma and Injury Severity Score

Introduction
============
Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].

Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi]. Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath]. However, the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made. A large population worldwide stands to gain from the improvement of trauma outcomes. These gains benefit not only the individual but also a broader population by reducing the societal and healthcare burden of disability. Reducing the economic burden of trauma is particularly relevant in light of the fact that low- to middle-income countries make up the majority of poor trauma outcomes.

The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences. M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review. In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal considering that it cannot be known whether a case was mismanaged prior to it being reviewed. This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo]. Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo]. The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours. The latter benefit is not insignificant considering the large caseload found in most urban trauma centres. The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines]. Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi]. Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi]. Another potential audit filter, that is already used at some clinics [@whoguidelines], is trauma severity scoring.

@@

@@

<!-- Potentially talk about other limitations of TRISS here? Feels somewhat unnecessary because the project isn't really about investigating every weakness of TRISS -->

In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit]. NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS. NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region. Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to the ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter]. NORMIT also provides a method to incorporate the GCS for intubated patients when calculating RTS. For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit]. NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit].

@@

As already mentioned, some clinics use TSS as audit filters to aid in identifying cases for M&M review. However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.

@@

There are far more TSS models than the two described above. These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is the location of this institution. It seems reasonable to begin by investigating these two models. The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.

@@

<!-- Also feels like I maybe have too many different sources cited for the intro, ~22 feels like overkill? -->

@@

<!-- Present key elements of study design -->
This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database (TCQD). The TCQD is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysed to estimate the association between the common TSS models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

The TCQD is an internal hospital register and registers the outcome of M&M conferences. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review. The database contains data on patients from 2014 to 2021.

@@

Participants include all patients registered in both the trauma registry and the TCQD at KUH that were not excluded by study-specific exclusion criteria. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->

@@

### Inclusion criteria for the TCQD

The TCQD contains patients from the trauma registry that were selected for M&M review. At KUH, all fatalities are automatically selected for M&M review. In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care. These nurses also use audit filters that automatically highlight cases with abnormal parameters. Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.

@@

<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. -->

@@

Since all eligible participants must exist in both the trauma registry and the TCQD, and the TCQD is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the TCQD. The TCQD includes around 2200 patients who were selected for review between 2014 and 2021. In order to establish the exact study size, the study-specific exclusion criteria need to first be applied. The exact study size has not yet been established at the time of writing this report.

@@

@@

All patient information used in this study is anonymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the KUH trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register on request. They also have the right to a copy of data that is stored about them on request.

@@

<!--

source("main.R")  "imports" the main script file of your project and run any code in it -->

>>>>

---
title: 'Association between trauma severity models and opportunities for improvement:
  A retrospective cohort study'
author: "Victoria Witthoff"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
    number_sections: true
bibliography: bibliography.bib
csl: vancouver.csl
fontsize: 11pt
linestretch: 1
always_allow_html: yes
---

@@

```{r main, echo=FALSE, message = FALSE}
  source("main.R") ## This "imports" the main script file of your project and run any code in it
```

Abbreviations
=============
* ASA-PS - American Society of Anesthesiology Physical Status
* ISS - Injury Severity Score
* KUH - Karolinska University Hospital
* M&M - Morbidity and mortality
* NISS - New Injury Severity Score
* NORMIT - Norwegian Survival Prediction Model in Trauma
* OFI - Opportunities for improvement
* Ps - Probability of survival
* RTS - Revised Trauma Score
* TRISS - Trauma and Injury Severity Score

Introduction
============
Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi]. Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath]. However, the implementation of QI programs differs widely between clinics [@tqipanglo], with smaller, non-specialised hospitals tending to have less organised programs [@whoguidelines].

@@

<!-- I know that a lot of the following details about TRISS/NORMIT are probably not really relevant to my aim, it's mainly here to fill out on length -->

@@

In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit]. NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS. NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region. Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter]. NORMIT also provides a method to incorporate the GCS for intubated patients when calculating RTS. For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit]. NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit].

The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences. M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The WHO themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review. In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal as it cannot be known whether a case was mismanaged prior to it being reviewed. This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo]. The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours. The latter benefit is not insignificant considering the large caseload found in many urban trauma centres. Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].

The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines]. Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi]. Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].

@@

Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring. However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.

@@

There are far more TSS models than the two described above. These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is where this institution is located. It seems reasonable to begin by investigating these two models. The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.

@@

<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? -->

@@

This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysed to estimate the association between the common TSS models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

The trauma care quality database is an internal hospital register and registers the outcome of M&M conferences. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review. The database contains data on patients from 2014 to 2021.

@@

Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->

@@

### Inclusion criteria for the trauma care quality database

<!-- 
``` {r, echo = FALSE,} 
  ofi_diagram() ## Flowchart of the process of arriving at OFI 
```
Removed this b/c it only works in html, diagrammersvg doesn't seem to be possible to install
-->

The trauma care quality database contains patients from the trauma registry that were selected for M&M review. At KUH, all fatalities are automatically selected for M&M review. In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care. These nurses also use audit filters that automatically highlight cases with abnormal parameters. Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.

@@

<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. Do I need to have the formulas? -->

@@

Since all eligible participants must exist in both the trauma registry and the trauma care quality database, and the trauma care quality database is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the trauma care quality database. The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021. In order to establish the exact study size, the study-specific exclusion criteria need to first be applied. The exact study size has not yet been established at the time of writing this report.

@@

<!-- BTW this is here because, according to me and what I see in the database, it seems like I'll have to calculate NORMIT/TRISS for everyone based on parameters in the database -->

@@

All patient information used in this study is anonymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the KUH trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register on request. They also have the right to a copy of data that is stored about them on request. <!-- https://meetings.lg.se/pdfFile/pdfFile.aspx?objectID=%2F7Iqgtt89nW2zhRjMSNZ%2Fw%3D%3D -->

@@
